Select-based random access \\ to variable-byte encodings

### 2020-04-14 gradu/msc_english_contents.tex
Adds subarray examples

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted.

@@

The $rank$ method is found superior when dataset contains mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based 
method performs better. Additionally, the proposed $select$ method offers fast access to the next or previous element.

@@

A single direct access with $select$ needs to calculate the length of the desired element. This is done by reading a 64-bit word from the continuation bit array, bit-shifting until the bit representing 
the first block is on the rightmost end, and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
and then counting trailing zeros. Once the word is shifted through, another word is read and process continues.

The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
When a new block level is reached for the first time during a subarray query, the index is fetched with $rank_1$ as normal, and the result of the query is stored. The next time the same level is reached, the
previously stored result is incremented by one and that value is used instead of issuing another $rank_1$ query. This way the number of $rank_1$ calls needed is equal to the number of $rank_1$ calls 
needed to decompress the largest number in the subarray.

 - examples / visualizaton

@@

\chapter{Conclusions Future work}

>>>>

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used a clever block 
reorganizing and a fast $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted \citep[e.g.][]{Kon17}.

@@

The $rank$ method is found superior when dataset contains mostly small integers. With larger integers, the decoding needs multiple $rank$ calls and the new $select$-based 
method performs better. Additionally, due to how the data is stored, the proposed $select$ method offers fast access to the next or previous element. This makes a subarray fetch 
very convenient and fast. A solution for subarray fetching and comparison is proposed in this thesis.

@@

\section{Subarray access with $select$}
A single direct access with $select$ needs to calculate the length of the desired element. This is done by reading a word from the continuation bit array, bit-shifting until the bit representing 
the first block is on the rightmost end, and then counting the trailing zeros in the word. Subsequent length calculations are done by bit-shifting the word equal to the length of the previous element 
and then counting trailing zeros. Once the word is shifted through, another word is read and process continues.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
[node distance=0pt,
 start chain = A going right,
 start chain = B going right,
 arrow/.style = {draw=#1,-{Stealth[]}, 
                shorten >=1mm, shorten <=1mm}, % styles of arrows
 arrow/.default = black,
 X/.style = {rectangle, draw,% styles of nodes in string (chain)
                minimum width=2ex, minimum height=5ex,
                outer sep=0pt, on chain},
 Y/.style = {rectangle, draw=none,% styles of nodes in string (chain)
                minimum width=0ex, minimum height=5ex,
                outer sep=0pt, on chain}
 ]

\node[X, join] (start) {...};
\node[X] {0};
\node[X] {1};
\node[Y] {};
\node[X] (byte1) {0};
\node[X] {1};
\node[X] {0};
\node[X] {1};
\node[X] {0};
\node[X] (select){1};
\node[X] {0};
\node[X] {0};
\node[Y] {};
\node[X] (byte2) {1};
\node[X] {1};
\node[X] {0};
\node[X] {0};
\node[X] {1};
\node[X] {0};
\node[X] {1};
\node[X] {0};
\node[Y] {};
\node[X] (byte3){0};
\node[X] {1};
\node[X] {0};
\node[X] {1};
\node[X] {1};
\node[X] {0};
\node[X] {1};
\node[X] {0};
\node[Y] {};
\node[X] (byte4){1 };
\node[X] {...};

\node (index)[above=2ex of select]{k-1th 1};
\node [below=1ex of byte1]{$B_1$};
\node [below=1ex of byte2]{$B_2$};
\node [below=1ex of byte3]{$B_3$};
\node [below=1ex of byte4]{$B_4$};
\draw[arrow] (index.south) to (select.north);
\end{tikzpicture}
\caption{Continuation bit array, the bits are grouped by the bytes they are in} \label{subarray}
\end{figure}

Figure~\ref{subarray} shows an example of a continuation bit array. The index of the 1 bit returned by the $select$ query issued for the starting index is marked, and the desired starting index 
$i$ = $select$($k-1$) +1 is the following bit on the right. The byte where $i$-th bit is located can be calculated by doing an integer division $i$/8. A 16-bit word $w$ is read from the byte location $B_1$.
It is beneficial to read as large word as possible, but 16-bits is used in this example for the sake of simplicity. Next the starting bit is shifted to the start of the word, the offset being 6 in this case. 
After the operation, $w$ contains bits \texttt{00110010 10000000}, least significant bit being on the left. The length of the current element is $l$ = 1 + \textsc{TrailingZeros}($w$). The number of 
trailing zeros can be efficiently calculated with GCC function \texttt{\_\_builtin\_ctz} or \texttt{bits::lo} from the 'SDSL-lite' library.

The length of the desired element $l$ and the location $i$ can now be used to decompress desired element. For the next element, $i$ is incremented by $l$ and $w$ is bit shifted $l$ bits and the process is 
repeated. When $w$ has no 1 bits, another word is read from the following bytes. A few extra steps are required, because the old $w$ might have continuation 0 bits left for the current word. The number of leftover 
zeros is added to the first length calculation after word switch. This number can be calculated either by adding bit shift amounts together and subtracting the result from word-length. Another option is to count 
leading zeros of the old $w$ before the first switch. In the example, one zero is left over from the byte $B_2$.

\section{Subarray access with $rank$}
The shown $rank$ method by Brisaboa et al. can also effectively decompress the next value with slight modifications. When fetching a longer subarray, the goal becomes to minimize $rank$ calls. 
When a new block level is reached for the first time during a subarray query, the index is fetched with $rank_1$ as normal, and the result of the query is stored. The next time the same level is reached, the
previously stored result is incremented by one and that value is used instead of issuing another $rank_1$ query. This way the number of $rank_1$ calls needed is equal to the number of $rank_1$ calls 
needed to decompress the largest number in the subarray.

 - examples / visualizaton for rank?

@@

\chapter{Conclusions and Future work}
